One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-10-15 doc/response.tex

We have attached a revised version of our manuscript ``A simple null model for inferences from network enrichment analysis'', and a diff
against our original manuscript. We have also answered the reviewer's comments
below.

In general, by reading the reviewers comments, we feel that we failed to deliver the purpose of the manuscript to the reviewer, who indeed appears to have misunderstood what we are aiming at with our manuscript. We have hence added clarifications to the introduction and the discussion. We hope that this will make the story even more interesting.

Yours Sincerely,\\[1.5cm]
Gustavo Jeuken and
Lukas K\"{a}ll
\\

\section*{Reviewer \#1}
\breview
Jeunken and K\"{a}ll describe a method for generating a null distribution for so called ``network enrichment analyses'', which are sometimes framed as ``crosstalk'' analyses. The only data presented is showing that the method generates accurate pvalues under the null.

As far as I can tell the work is technically correct. Their statement of the null hypothesis they are testing is sufficiently precise. I agree with the authors that their conceptualization of the problem is simpler than methods like BinoX. But I disagree that the interpretation can be assumed to be meaningful, and the authors provide no evidence.
\ereview
The reviewer points out that the manuscript only discusses the calibration of $p$ values, and never actually demonstrates the utility of such $p$ values. The reviewer even does not find the findings of the manuscript meaningful. While it is worth noting that  according to the Criteria For Publication PLOS ONE, (\url{https://journals.plos.org/plosone/s/criteria-for-publication}) perceived meaningfulness is not a publication criterion for a manuscript, lack of meaning is obviously an undesirable property of any scientific report. We have carefully rewritten the introduction to make the readers get a sense of the underlying conflict driving our manuscript. However, we tried to not to bee too direct when writing that existing methods for network enrichment analysis (NEA) produce grossly inflated statistics. We instead have tried to, in a constructive manner, described a method for deriving statistics that is natural from a users perspective. This model could be integrated into any of the currently used  models for NEA. It is an open question if one could demonstrate improved sensitivity with NEA over over-representation analysis in such implementations.

@@

\breview
They claim that BinoX yields very inaccurate p-values. If the results yielded by the proposed method are basically the same as BinoX (in terms of ranks), then there is no major problem. But the authors state that they don’t want to get into evaluating ``performance in terms of sensitivity''. I find this disingenuous; they have to at least show how their ranked results compare to BinoX on real data and do some kind of benchmarking. This is important because I believe there is a difference in the null hypothesis stated in the current work and that used by BinoX that makes the comparison presented here unfair -- they are not testing the same thing.
\ereview

We fully agree with the reviewer that the null of BinoX is different that GeneSetDP. In fact that is the purpose of our manuscript. One reason why no sensitivity test is made is that there is no way to compete in sensitivity against a method with inflated statistics. If the yardstick is flawed no comparison can be made. Here we establish a fair yardstick to compare against. In our manuscript we are not trying to sell a new method for NEA, we are merely trying to convince the field that they should report statistics that makes sense to a user.
Furthermore, there is nothing to benchmark but the statistics itself. We use the same scoring system as BinoX, i.e. number of links.

@@

\breview
The crux is that the method proposed here focuses on the properties of the query genes; randomly sampling query genes makes the proposed method a test of whether the query genes are unusual, not necessarily specifically with respect to a particular pathway. For example, if the query contains a hub gene (assuming hubs are rare) then it might be found ``significantly'' connected to many pathways. The ``feature unlikely to occur by chance'' is that the query contains a hub gene, not that it connects to many pathways. To state it in an extreme way, in this situation biologist should skip the pathway analysis and just take note of the fact that the query includes a hub. Using the authors’ method without this insight could lead to very misleading conclusions.

\medskip

As I understand it, BinoX tries to account for these difficulties by modeling realizations of the network that retain key topological properties like node degree distribution while holding the query constant (in effect). The current authors criticize the BinoX approach, but ignore the problem entirely. A closer (but not very close) approximation of the BinoX approach might be, for the Monte Carlo method proposed here, to bias the sampling of $Q’$ from $G$ such that the distribution of the node degrees of the genes in $Q’$ resembles the distribution of the actual query set of interest $Q$. In any case, I believe that the BinoX approach is not directly comparable. The relationship needs to be discussed and some demonstration of the method yielding useful and specific results is needed. If the authors cannot do this then they must remove the discussion of BinoX. Without that, and without any benchmarking, the paper is essentially about an algorithm to compute $N(s)$, with no biological content or relevance.

@@

\ereview

Here, we respectfully disagree with the reviewer. The reviewer's perspective do not reflect the actual use case. The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in respect to a pathway database.  In this question, the introduction of a  network is not relevant more than in that it might shift the score distribution against pathways. A network may or may not help to increase the sensitivity of ORA methods, but it is better left out of the null hypothesis.  A user should not have to worry about the connectivity of individual genes. Just as the reviewer points out, there are certain hub genes that are connected to more pathways than others through any network. This will be reflected by the score distribution produced by our described methods. If the query set contains such a gene, the query is more likely to be significant against a pathway, however, that is the desired property of the methods. Conversely, any query that contain one or more genes that are not connected to any pathways is less likely to be significantly enriched. This is the nature of this type of analysis.

However, a perhaps more problematic fact is that the methods in this manuscript, just in all NEA methods, test every pathway independently for significance, and do not evaluate correlation between different pathways.

### 2018-10-16 doc/genesetdp.tex
Edited together.
small text changes

% Use the asterisk to denote corresponding authorship and provide email address in note below.
*lukas.kall@scilifelab.se
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
A prevailing technique to infer function from lists of identifications, from molecular biological high-throughput experiments, is over-representation analysis, where the identifications are compared to predefined sets of related genes often referred to as pathways. As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks. While the terminology varies in the literature, we will here refer to such methods as Network Enrichment Analysis (NEA). Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations. Here we instead have designed one dynamic programming algorithm that calculates the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences. We also implemented a random sampling method, carrying out the same task. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX, while also providing statistics that are easier to interpret.

@@

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step, they must follow constraints that depend on previous link assignments.
While this results in a network that is both randomized and contains some desired topological features preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative. In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.  And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose of the statistical test.

@@

We note that the calibration of GeneSetDP is slightly conservative, i.e. the calculated $p$ values are larger than expected. Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are lower than expected, even when not taking into account the significance by depletion.

@@

Here we have implemented two methods, GeneSetDP and GeneSetMC, to calculate unbiased $p$~values for inferences from NEA.
Instead of testing our methods' performance in terms of sensitivity, we here instead chose to test the method in terms of the methods statistical accuracy. We argue that more studies should make a point at demonstrating that their methods are well-calibrated, as it is a prerequisite for measuring performance at least when the ground truth of biological effect is not known.
% With this in mind, we demonstrated that both our methods gave a better calibration than the popular BinoX method, by random sampling of subsets of a genome.

@@

We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$: ``There are not more links between the query and pathway gene sets than expected by chance''. Previous implementations of NEA in the literature use network randomization methods to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use for calculating accurate statistics. Generally, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters. 
In this line, network randomization methods would be better suited for studying the relevance of the links in the network than the query set itself.

In effect, NEA is an attempt to increase the sensitivity of ORA. The standard user of NEA would typically have identified a set of query genes which she wants to evaluate for over-representation in respect to a pathway database. This goal is the same as for users of ORA. Hence, the null model should be of the same nature for NEA and ORA. GeneSetDP and GeneSetMC are two algorithms that prove that one in NEA can test the uniqueness of the query set in respect to a pathway, just as done in ORA.

>>>>

% Use the asterisk to denote corresponding authorship and provide email address in note below.
*lukas.kall@scilifelab.se
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
A prevailing technique to infer function from lists of identifications, from molecular biological high-throughput experiments, is over-representation analysis, where the identifications are compared to predefined sets of related genes often referred to as pathways. As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks. While the terminology varies in the literature, we will here refer to such methods as Network Enrichment Analysis (NEA). Traditionally, the significance of inferences from NEA has been assigned using a null model that derives from a randomization of the network. Here we instead argue for a null model that more direcly relates to the set of genes being studied, and have designed one dynamic programming algorithm that calculates the score distribution of NEA scores that makes it possible to assign unbiased mid $p$~values to inferences. We also implemented a random sampling method, carrying out the same task. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX, while also providing statistics that are easier to interpret.

@@

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative. In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.  And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose of the statistical test.

@@

We note that the calibration of GeneSetDP is slightly conservative, i.e. the calculated $p$ values are larger than expected. Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are lower than expected, even when not taking into account significance by depletion.

@@

Here we have implemented two methods, GeneSetDP and GeneSetMC, to calculate unbiased $p$~values for inferences from NEA.
Instead of testing our methods' performance in terms of sensitivity, we here instead chose to test the method in terms of the methods statistical accuracy. We argue that more studies should make a point at demonstrating that their methods are well-calibrated, as it is a prerequisite for measuring performance, most importantly when the ground truth of biological effect is not known.
% With this in mind, we demonstrated that both our methods gave a better calibration than the popular BinoX method, by random sampling of subsets of a genome.

@@

We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$: ``There are not more links between the query and pathway gene sets than expected by chance''. Previous implementations of NEA in the literature use network randomization methods to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are very heuristic in their nature, and hence are difficult to use for calculating accurate statistics. Generally, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.

Furthermore, network randomization methods focus on significance of the links in the network instead of the query set itself. In effect, NEA is an attempt to increase the sensitivity of ORA. The standard user of NEA would typically have identified a set of query genes which she wants to evaluate for over-representation in respect to a pathway database. This goal is the same as for users of ORA. Hence, the null model should be of the same nature for NEA and ORA. GeneSetDP and GeneSetMC are two algorithms that prove that one in NEA can test the uniqueness of the query set in respect to a pathway, just as done in ORA.
